---
title: "P3 Assignment Notebook"
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

Import all data sets given

```{r}
setwd("/Users/ganeshharugeri/Documents/R_workspace/input/P3")
print(c("Working Directory:",getwd()))

filename1 <- paste(c(getwd(),"/flights.csv"),collapse="")
filename2 <- paste(c(getwd(),"/airlines.csv"),collapse="")
filename3 <- paste(c(getwd(),"/airports.csv"),collapse="")

flights <- read.table(filename1,sep = ",",header = 1)
airlines <- read.table(filename2,sep = ",",header = 1)
airports <- read.table(filename3,sep = ",",header = 1)

```

Data loading: Solving the Data loading issue due to sererator difference (, ; mismatch) and then reloading.

```{r}
#rm(flights)
#rm(airlines)
#rm(airports)
```


Understanding the structure of the given datasets

```{r}

str(flights)
str(airlines)
str(airports)
``` 

Data preperation: Deleted incomplete columns, having more than 80% of missing values

```{r}
flights <- subset(flights, select = -c(27:31) )
```


Data Exploration: See the data at a glance and ther samples
```{r}
print(c("Flights Sample data", head(flights)))
print(c("Airlines Sample data",head(airlines)))
print(c("Airports Sample data",head(airports)))
```



Data Exploration: Summary analysis of the given data sets
```{r}
print(c("Flights Summary", summary(flights)))
print(c("Airlines Summary",summary(airlines)))
print(c("Airports Summary",summary(airports)))
```


Data Transformation: Enabling data type conversions to increase effciency of the data readability. 

```{r}
print (c("YEAR:", levels (factor (flights$YEAR))))
print (c("MONTH:", levels (factor (flights$MONTH))))
print (c("DAY:", levels (factor (flights$DAY))))
print (c("DAY_OF_WEEK:", levels (factor (flights$DAY_OF_WEEK))))

print (c("DIVERTED:", levels (factor (flights$DIVERTED))))
print (c("CANCELLED:", levels (factor (flights$CANCELLED))))

# Data converted as they had only two levels 
flights$DIVERTED <- as.factor(flights$DIVERTED)
flights$CANCELLED <- as.factor(flights$CANCELLED)
flights$DAY_OF_WEEK <- as.factor(flights$DAY_OF_WEEK)
flights$MONTH <- as.factor(flights$MONTH)
flights$YEAR <- as.factor(flights$YEAR)

typeof(flights$CANCELLED)
```


Data Representation & flights distribution per month plot

```{r}
table(flights$MONTH)
library(ggplot2)
ggplot(flights, aes(x=MONTH, fill=MONTH))+geom_bar()
#ggplot(flights, aes(x=Day, fill=DayofMonth))+geom_bar()

#Another method
#x<-table(flights$MONTH)
#barplot(x)
```


I tried to remove the rows with incomplete details as the scope of this work is not buid a perfect predicting model.

```{r}
flights<-flights[complete.cases(flights), ]
na.omit(flights)
```



Data Exploration: Market shares by Air carriers

```{r}
flights <- subset(flights, !(is.na(flights$AIR_TIME)))
na_percentages <- sapply(flights, function(x) sum(is.na(x))) / nrow(flights) * 100
sort(format(round(na_percentages, 4), nsmall = 4))
```


Data Exploration: Delay analysis per airline carrier

```{r}
airline.avg.delay <- aggregate(flights$ARRIVAL_DELAY, by=list(flights$AIRLINE), mean, na.rm=T)
names(airline.avg.delay) <- c("AirlineCode", "Mean.Arrival.Delay")
airline.avg.delay <- merge(airline.avg.delay, airlines, by.x="AirlineCode", by.y="IATA_CODE", all.x=TRUE)

airline.avg.delay <- airline.avg.delay[order(airline.avg.delay$Mean.Arrival.Delay), ]
airline.avg.delay <- airline.avg.delay[ ,c(3,1,2)]

airline.avg.delay


barplot(airline.avg.delay$Mean.Arrival.Delay,
        names.arg=airline.avg.delay$AirlineCode,
        col="grey",
        main="Mean Arrival Delay by Airline", 
        xlab="Airline Code",
        ylab="Mean Arrival Delay")
```




Data Splitting: data is divided to training and testing datasets:

```{r}

#library(caret)
#ind <- sample(2, nrow(flights), replace = TRUE, prob=c(0.7, 0.3))

#Too high memory required for the random arrangements
#trainset = flights[ind == 1,]
#testset = flights[ind == 2,]

rm(flights_train)
rm(testData)
set.seed(120)
trainset = flights[1:4073355,]
testset = flights[4073355:5819079,]

```


Create categorical columns for the trainset.

```{r}
#na.omit(trainset)
#na.omit(flights)

trainset$delayed = trainset$ARRIVAL_DELAY > 0
trainset$on_time = trainset$ARRIVAL_DELAY == 0
trainset$ahead_of_time = trainset$ARRIVAL_DELAY < 0

#flights_train <- colMeans(flights[,-23])
#flights_train = subset(flights, select = -c(23))
#str(flights_train)
#testData = flights_train[1:400,]

```


Installation and initiation Neural net package
Neural net for the newly created categorical attribues
```{r}
# Too big
library(neuralnet)
nn_result = neuralnet(delayed + on_time + ahead_of_time ~ SCHEDULED_DEPARTURE + DEPARTURE_DELAY + TAXI_OUT + TAXI_IN + SCHEDULED_ARRIVAL + ARRIVAL_TIME, trainset, hidden=1)

#nn_result

```

 

Result matrix - Use it to predict flight delay

#Predict future
```{r}
nn_result$result.matrix
#nn_pred <- compute(nn_result, testset)
#Run them through the neural network

```

Plot the Neural network
```{r}
plot(nn_result)
```

